D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-02-14 ootavacuous.tex
Change "partially volatile" to "quasi volatile"
At the risk of seeming over-eager, I'm going to change the name we use
from "partially volatile" to "quasi volatile" now.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

	\co{volatile} is a hint to the implementation to avoid aggressive
	optimization involving the object because the value of the object
	might be changed by means undetectable by an implementation.
\end{quote}
Following the folklore, we interpret these comments to mean that
volatile accesses should be treated as a form of I/O,
implying that volatile loads can return unpredictable values.
They don't necessarily read from stores (in contrast to non-volatile loads,
which must return the value of the store they read from).
As a corollary, the rfe relation does not apply to volatile loads
and hence the accesses in an OOTA cycle must not be volatile.
We also understand that a compiler may not
invent, omit, merge, or reorder volatile accesses,
in a sense to be discussed in
section~\ref{sec:Volatile and Partially Volatile Accesses} below.

@@

A noteworthy aspect of this transformation is that it invents an atomic load:
The original form of the code reads \co{x} only once,
whereas the transformed code reads it twice.
Therefore we can rule out transformations like this one by insisting
the compiler not invent atomic loads.
In fact, we will require that atomic accesses be treated as
``partially volatile'', in that the compiler is allowed to omit,
merge, or reorder them but not invent them.

@@

\subsection{Volatile and ``Partially Volatile'' Accesses}
\label{sec:Volatile and Partially Volatile Accesses}

@@

For accesses to partially volatile objects, we remove some of the
requirements and add a new one.
We leave out the last two items; compilers are allowed to merge or
omit accesses to partially volatile objects.
And the requirement for order preservation is weakened; it applies
only to pairs of accesses to the same object.
Accesses to different objects may be reordered relative to each other.
The new requirement is that the map must preserve the rf relation.
That is, if load $R$ in $A$ reads from store $W$ and $R$ is not
omitted then the instruction it maps to in $H$ must read from the
instruction that $W$ maps to,
and consequently $W$ must not be omitted either.

@@

With the contents of the last few sections under our belts, we can
formulate the Fundamental Property that we would like all semantic
dependencies to satisfy.
Note that this formulation makes sense only for implementations in
which all atomic objects are treated as partially volatile.
\begin{quote}
Let $W$ be a store which, in some abstract execution $A$, semantically
depends on loads $\{R_0$, \ldots,~$R_n\}$, and suppose that $W$ is not
omitted in some hardware execution $H$ realizing $A$.
Then for some $i$, load $R_i$ is not omitted and the hardware
instruction it maps to is ordered before the instruction $W$ maps to.
\end{quote}
It's quite straightforward to show that for any implementation in which
semantic dependencies satisfy the Fundamental Property, no OOTA cycle
has a nontrivial realization.

>>>>

	\co{volatile} is a hint to the implementation to avoid aggressive
	optimization involving the object because the value of the object
	might be changed by means undetectable by an implementation.
\end{quote}
Following the folklore, we interpret these comments to mean that
volatile accesses should be treated as a form of I/O,
implying that volatile loads can return unpredictable values.
They don't necessarily read from stores (in contrast to non-volatile loads,
which must return the value of the store they read from).
As a corollary, the rfe relation does not apply to volatile loads
and hence the accesses in an OOTA cycle must not be volatile.
We also understand that a compiler may not
invent, omit, merge, or reorder volatile accesses,
in a sense to be discussed in
section~\ref{sec:Volatile and Quasi Volatile Accesses} below.

@@

A noteworthy aspect of this transformation is that it invents an atomic load:
The original form of the code reads \co{x} only once,
whereas the transformed code reads it twice.
Therefore we can rule out transformations like this one by insisting
the compiler not invent atomic loads.
In fact, we will require that atomic accesses be treated as
``quasi volatile'', in that the compiler is allowed to omit,
merge, or reorder them but not invent them.

@@

\subsection{Volatile and ``Quasi Volatile'' Accesses}
\label{sec:Volatile and Quasi Volatile Accesses}

@@

For accesses to quasi volatile objects, we remove some of the
requirements and add a new one.
We leave out the last two items; compilers are allowed to merge or
omit accesses to quasi volatile objects.
And the requirement for order preservation is weakened; it applies
only to pairs of accesses to the same object.
Accesses to different objects may be reordered relative to each other.
The new requirement is that the map must preserve the rf relation.
That is, if load $R$ in $A$ reads from store $W$ and $R$ is not
omitted then the instruction it maps to in $H$ must read from the
instruction that $W$ maps to,
and consequently $W$ must not be omitted either.

@@

With the contents of the last few sections under our belts, we can
formulate the Fundamental Property that we would like all semantic
dependencies to satisfy.
Note that this formulation makes sense only for implementations in
which all atomic objects are treated as quasi volatile.
\begin{quote}
Let $W$ be a store which, in some abstract execution $A$, semantically
depends on loads $\{R_0$, \ldots,~$R_n\}$, and suppose that $W$ is not
omitted in some hardware execution $H$ realizing $A$.
Then for some $i$, load $R_i$ is not omitted and the hardware
instruction it maps to is ordered before the instruction $W$ maps to.
\end{quote}
It's quite straightforward to show that for any implementation in which
semantic dependencies satisfy the Fundamental Property, no OOTA cycle
has a nontrivial realization.
